---
title: "POL501 - Problem Set 3"
author: "Answers to Questions"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    keep_tex: true  # Optional, keeps the .tex file for inspection
fontsize: 11pt
geometry: "left=1in,right=1in,top=1in,bottom=1in"
header-includes: 
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines, breakanywhere, commandchars=\\\{\}}
---

\newpage


```{r setup, include=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(haven)
library(tidyverse)
library(formatR)
```


# Grading Criteria

- There will be **four problem sets** throughout the semester, which together account for **25% of the final course grade**.
- The total possible score for these problem sets is **25 out of 100 points** (with 100 points being the maximum course score).
- This particular problem set has a maximum score of **8 points**.
- **Scoring Breakdown**:
  * Each sub-letter in Question 1 is worth **1 point** (Total of 5 points)
  * Each sub-letter in Question 2 is worth **0.5 points** (Total of 3 points)
  
- **Grading Guidelines**:
  * Full credit will be awarded for answers that closely match the provided solutions.
  * Partial credit will be given for incomplete or partially incorrect answers/justifications.
  * **0 points** will be awarded for missing answers, answers with no justification, or entirely incorrect responses.
  * Submissions without the RMD file will have point deductions.

# Question 1 - Solutions

## Answer - Question (1.a)

```{r 1a, echo=TRUE}
```

## Answer - Question (1.b)

```{r 1b, echo=TRUE}
```

## Answer - Question (1.c)

```{r 1c, echo=TRUE}
```

## Answer - Question (1.d)

```{r 1d, echo=TRUE}
```

## Answer - Question (1.e)

```{r 1e, echo=TRUE}
```


# Question 2 - Solutions

## Answer - Question (2.a)

```{r 2a, echo=TRUE}
```

## Answer - Question (2.b)

```{r 2b, echo=TRUE}
```

## Answer - Question (2.c)

```{r 2c, echo=TRUE}
```

## Answer - Question (2.d)

```{r 2d, echo=TRUE}
```

## Answer - Question (2.e)

```{r 2e, echo=TRUE}
```

## Answer - Question (2.f)

```{r 2f, echo=TRUE}
```



