---
title: "Introduction to dplyr"
author: "POL-501"
output: 
  word_document:
    toc: true
    toc_depth: 2
    number_sections: true
date: "2024-09-16"
---

# Introduction

This document will introduce you to the `dplyr` package in R, which is part of the tidyverse collection of packages designed for data science. The `dplyr` package is particularly powerful for data manipulation, providing an intuitive syntax and a set of versatile functions for managing data frames.

To illustrate the capabilities of `dplyr`, we will use data from Pew's American Trends Panel (ATP) Wave 117.

We have obtained the variables' coding (what each numeric value represents) from the PDF files 'ATP W117 questionnaire' and 'Codebook-and-instructions-for-working-with-ATP-data'

### Loading Necessary Libraries

Before we begin, ensure you have the required packages installed and loaded. If not, you can install them using `install.packages()`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# BE SURE TO INCLUDE YOUR FILE DIRECTORY HERE AND BELOW (using the 'setwd' function)
knitr::opts_knit$set(root.dir = "C:/Users/Ignacio/Dropbox/PhD SBU/06_Teaching/00_POL-501/Problem Sets/PS-1/W117_Nov22")
```

```{r load-libraries}
# Install libraries if necessary
# install.packages("haven")
# install.packages("dplyr")

# Load libraries
library(haven)  
library(dplyr)  
```

### Setting the Working Directory

Set the working directory to the folder where your data file is located. Adjust the path accordingly.

```{r set-working-directory}
# Set the working directory
setwd('C:\\Users\\Ignacio\\Dropbox\\PhD SBU\\06_Teaching\\00_POL-501\\Problem Sets\\PS-1\\W117_Nov22\\')
```

### Loading the Dataset

We'll load the dataset using the `read_spss()` function from the `haven` package.

```{r load-dataset}
# Load the Pew's ATP Wave 117 dataset
load_savfile <- read_spss("ATP W117.sav")
```

### Subsetting Relevant Columns

We can use the `select()` function from `dplyr` to pick specific columns from the dataset.

```{r subset-columns}
# Subset relevant columns
df_subset <- load_savfile %>%
  select(QKEY, VOTED_ATPCONG_W117, VTCOUNT_OWN_W117, ATTENDPERSON2_W117,
         F_AGECAT, F_GENDER, F_EDUCCAT)
# View the structure of the subset data
str(df_subset)
```

## Recoding Categorical Variables

### Recoding Voting Status

We will use the `mutate()` and `factor()` functions to recode the `VOTED_ATPCONG_W117` variable.

```{r recode-voting-status}
# Recode VOTED_ATPCONG_W117 (Voting status)
df_subset <- df_subset %>%
  mutate(VOTED_ATPCONG_W117 = factor(VOTED_ATPCONG_W117,
                                      levels = c(1, 2, 3),
                                      labels = c("Did not vote", 
                                                 "Planned to vote but couldn’t", 
                                                 "Definitely voted")))
```

### Recoding Confidence in Vote Count

Similarly, we recode `VTCOUNT_OWN_W117`.

```{r recode-vote-confidence}
# Recode VTCOUNT_OWN_W117 (Confidence in vote count)
df_subset <- df_subset %>%
  mutate(VTCOUNT_OWN_W117 = factor(VTCOUNT_OWN_W117,
                                   levels = c(1, 2, 3, 4),
                                   labels = c("Very confident", 
                                              "Somewhat confident", 
                                              "Not too confident", 
                                              "Not at all confident")))
```

### Recoding Religious Service Attendance

We now recode `ATTENDPERSON2_W117`.

```{r recode-religious-attendance}
# Recode ATTENDPERSON2_W117 (Religious service attendance frequency)
df_subset <- df_subset %>%
  mutate(ATTENDPERSON2_W117 = factor(ATTENDPERSON2_W117,
                                     levels = c(1, 2, 3, 4, 5, 6),
                                     labels = c("More than once a week", 
                                                "Once a week", 
                                                "Once or twice a month", 
                                                "A few times a year", 
                                                "Seldom", 
                                                "Never")))
```

## Recoding Demographic Variables

### Recoding Age Categories

Recode the `F_AGECAT` variable into readable age ranges.

```{r recode-age-categories}
# Recode F_AGECAT (Age categories)
df_subset <- df_subset %>%
  mutate(F_AGECAT = factor(F_AGECAT,
                           levels = c(1, 2, 3, 4, 99),
                           labels = c("18-29", 
                                      "30-49", 
                                      "50-64", 
                                      "65+", 
                                      "Refused")))
```

### Recoding Gender Categories

Recode the `F_GENDER` variable.

```{r recode-gender-categories}
# Recode F_GENDER (Gender categories)
df_subset <- df_subset %>%
  mutate(F_GENDER = factor(F_GENDER,
                           levels = c(1, 2, 3, 99),
                           labels = c("Man", 
                                      "Woman", 
                                      "Other",
                                      "Refused")))
```

### Recoding Educational Attainment

Finally, recode the `F_EDUCCAT` variable.

```{r recode-education-categories}
# Recode F_EDUCCAT (Educational attainment categories)
df_subset <- df_subset %>%
  mutate(F_EDUCCAT = factor(F_EDUCCAT,
                            levels = c(1, 2, 3, 99),
                            labels = c("College graduate+", 
                                       "Some college", 
                                       "H.S. graduate or less", 
                                       "Refused")))
```

## Confirming the Structure

Check the structure and ensure the new labels have been applied correctly.

```{r check-structure}
# Confirm the structure and labels are correctly applied
str(df_subset)
```

## Viewing the First Few Rows

Optionally, we can view the first few rows to verify the recoding worked as expected.

```{r view-first-rows}
# View the first few rows to ensure everything looks correct
head(df_subset)
```

With this, you have successfully recoded categorical variables and formatted demographic variables using `dplyr`. Understanding these basic yet powerful functions will enable you to manipulate your datasets more effectively in R. This concludes our introduction to `dplyr`. 


# Introduction to dplyr

`dplyr` is a powerful package for data manipulation in R, providing a range of functions (often referred to as "verbs") to simplify and streamline data analysis tasks. Key verbs in `dplyr` include:

- `select()`: Pick columns by name
- `filter()`: Pick rows by their values
- `mutate()`: Create or transform variables
- `summarize()`: Reduce multiple values down to a single summary
- `arrange()`: Reorder rows
- `group_by()`: Group data by one or more variables

We also make extensive use of the pipe operator (`%>%`), which allows us to chain together multiple operations in a readable, functional style.

Note that the following is the structure of each operation:

```r
resulting_df_name <- df_used %>%
  verb1 %>%
  verb2 %>%
  verb3
```

Here are examples of using these verbs with the `df_subset` data.

### Select Columns

To select specific columns, we use the `select()` function:

```{r select-columns}
# Select specific columns
selected_data <- df_subset %>%
  select(QKEY, VOTED_ATPCONG_W117, F_AGECAT, F_GENDER, F_EDUCCAT)

# View the selected columns
head(selected_data)
```

### Filter Rows

To filter rows based on a condition, we use the `filter()` function:

```{r filter-rows}
# Filter data for respondents who definitely voted
voted_data <- df_subset %>%
  filter(VOTED_ATPCONG_W117 == "Definitely voted")

# View the filtered data
head(voted_data)
```

### Mutate Columns

To create new variables or transform existing ones, we use the `mutate()` function:

```{r mutate-columns}
# Create a new variable "Age_Group" based on F_AGECAT
df_subset <- df_subset %>%
  mutate(Age_Group = case_when(
    F_AGECAT == "18-29" ~ "Young Adult",
    F_AGECAT == "30-49" ~ "Adult",
    F_AGECAT == "50-64" ~ "Middle Aged",
    F_AGECAT == "65+" ~ "Senior"
  ))

# View the modified dataset
head(df_subset)
```


In R, we use `==` for comparison in conditional statements to check if two values are equal, whereas `=` is used for assigning values to variables. In the `case_when()` function, `==` is necessary because we are evaluating conditions (e.g., checking if a variable equals a specific value) rather than assigning values. The tilde `~` is used in `case_when()` to link each condition with its corresponding output value. It essentially means "if this condition is true, then return this value." Therefore, `==` checks the condition, and `~` specifies the result to be assigned when the condition holds true.

```{r view-result}
# Create a table with the absolute frequency of each age group
tab_Age <- as.data.frame(table(df_subset$Age_Group))  # Convert the frequency table to a data frame (use the function: as.data.frame())

# Rename the columns of the tab_Age dataframe
colnames(tab_Age) <- c("Age Group", "Freq")

# Print the table
print(tab_Age)
```

```{r bar-plot}
# Load ggplot2 for plotting
library(ggplot2)

# Create a bar plot using ggplot2
ggplot(tab_Age, aes(x = `Age Group`, y = Freq)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of Age Groups", x = "Age Group", y = "Frequency") +
  theme_minimal()

```

In `ggplot2`, the structure of a plot is built in layers, starting with the base `ggplot()` function, which initializes the plot and specifies the data and aesthetic mappings. In this example, `ggplot(tab_Age, aes(x = Age Group, y = Freq))` tells R that the data comes from the `tab_Age` dataframe and maps the "Age Group" variable to the x-axis and "Freq" (frequency) to the y-axis. After initializing the plot, the `geom_bar(stat = "identity")` layer is added to create the bars. Here, `stat = "identity"` ensures that the height of each bar corresponds to the actual frequency values in the dataset. Additional layers like `labs()` are used to label the plot, providing a title and axes labels. Finally, `theme_minimal()` applies a clean, minimal theme to enhance the visual appeal of the plot. Each of these components builds upon the last, allowing for flexible, modular creation of complex visualizations.

### Summarize Data

To summarize data, often in combination with `group_by()`, we use `summarize()`:

```{r summarize-data}
# Summarize data to count the number of respondents in each age group
age_summary <- df_subset %>%
  group_by(Age_Group) %>%
  summarize(Count = n())

# View the summarized data
age_summary
```

In this code, the pipe operator (`%>%`) is used to chain multiple `dplyr` functions in a readable sequence. The `group_by(Age_Group)` function groups the dataset (`df_subset`) by the `Age_Group` column, meaning that subsequent operations will be performed separately for each unique `Age_Group`. The `summarize()` function then creates a summary for each group, in this case calculating the number of rows (or observations) in each `Age_Group` using `n()`, which counts the number of rows in each group. The result is stored in a new column called `Count`, and the resulting dataset `age_summary` will have one row for each unique `Age_Group`, with a corresponding count of observations in that group.

### Arrange Rows

To reorder rows, we use the `arrange()` function:

```{r arrange-rows}
# Arrange data by age group
arranged_data <- df_subset %>%
  arrange(F_AGECAT)

# View the arranged data
head(arranged_data)
```

### Group By

To group data by one or more variables before summarizing, we use `group_by()`:

```{r group-by}
# Group data by gender and summarize age categories
gender_age_summary <- df_subset %>%
  group_by(F_GENDER, F_AGECAT) %>%
  summarize(Count = n())

# View the grouped and summarized data
gender_age_summary
```

### Example Analysis

Now, let's perform a more comprehensive example analysis where we combine multiple `dplyr` functions. Suppose we want to compare voting confidence based on different levels of education.

```{r example-analysis}
# Compare voting confidence based on education levels
edu_voting_confidence <- df_subset %>%
  group_by(F_EDUCCAT, VTCOUNT_OWN_W117) %>%
  summarize(Count = n()) %>%
  arrange(F_EDUCCAT, VTCOUNT_OWN_W117)

# View the analysis results
edu_voting_confidence
```

This pipeline will group the data by `F_EDUCCAT` (education levels) and `VTCOUNT_OWN_W117` (voting confidence), summarize the counts in each group, and then arrange the results for easier interpretation.

By understanding these key functions and how to use pipes to combine them, you'll be able to manipulate and analyze your data more effectively. This concludes our comprehensive introduction to `dplyr`. Explore these functions further to enhance your data analysis skills in R.

