---
title: "Handling Non-Applicable Survey Values"
author: "POL 501"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 2
geometry: "left=1in,right=1in,top=1in,bottom=1in"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Defining Non-Applicable Values in the Survey Context**

In survey data, **non-applicable values** are responses that do not provide meaningful information for analysis. These values typically represent cases where a question was skipped, the respondent didn't understand or chose not to answer, or the response is irrelevant due to the survey's flow logic.

Here are common types of non-applicable values in a survey context:

---

#### **1. Skipped Responses**
   - Represented by negative integers or large numbers.
   - Examples:
     - `-1`: Legitimate skip (e.g., the question did not apply to the respondent due to routing logic).
     - `-7`: Refusal to answer.
     - `-9`: Missing data for unknown reasons.

#### **2. Placeholder Codes for Missing or Irrelevant Responses**
   - Large numbers are often used as placeholders.
   - Examples:
     - `97`: Not applicable.
     - `98`: Don’t know.
     - `99`: Prefer not to say.

#### **3. Irrelevant Responses**
   - These may include codes used for respondents whose answers are logically unrelated to the question.
   - For example, a respondent who answered "No" to a question about being employed may skip a question about job satisfaction.

---

### **Why Transform Non-Applicable Values to `NA`?**
- Non-applicable values are **not numerical data** but placeholders.
- Leaving them as-is can distort analysis:
  - Inflates means or sums.
  - Causes errors in functions like regression or ANOVA.
- Replacing with `NA` allows R to ignore these values (e.g., using `na.rm = TRUE`).

### **Typical Handling in R**
1. Identify the non-applicable values specific to the survey (e.g., `-1`, `97`, etc.).
2. Replace them with `NA` for compatibility with analytical functions.


## Mock Dataset

We simulate a survey dataset with non-applicable values such as `-1`, `-7`, `97`, `98`, and `99`.

```{r mock-data}
library(dplyr)

# Create a mock dataset
set.seed(123)
survey_data <- data.frame(
  respondent_id = 1:20,
  support_abortion = sample(c(1, 2, 3, 5, -1, -7, 98), 20, replace = TRUE),
  religion_importance = sample(c(1, 2, 3, 4, 5, 99), 20, replace = TRUE),
  gender = sample(c("Male", "Female"), 20, replace = TRUE),
  vote_intent = sample(c("Democrat", "Republican", "Independent", "Other", NA), 20, replace = TRUE)
)

# View unique values in columns
table(survey_data$support_abortion)
table(survey_data$religion_importance)
```

## Replace Non-Applicable Values with NA

We transform non-applicable values (`-1`, `-7`, `97`, `98`, `99`) into `NA`.

```{r replace-na}
# Define non-applicable values
non_applicable <- c(-1, -7, 97, 98, 99)

# Replace non-applicable values with NA
survey_data_clean <- survey_data %>%
  mutate(
    support_abortion = ifelse(support_abortion %in% non_applicable, NA, support_abortion),
    religion_importance = ifelse(religion_importance %in% non_applicable, NA, religion_importance)
  )

# View cleaned data
table(is.na(survey_data_clean$support_abortion))
table(is.na(survey_data_clean$religion_importance))
```

## Handling NA in Analysis

With non-applicable values converted to `NA`, functions like `mean()` and `sum()` can now exclude them using `na.rm = TRUE`.

```{r handle-na}
# Example: Mean of support_abortion
mean_support_abortion <- mean(survey_data_clean$support_abortion, na.rm = TRUE)
mean_support_abortion

# Example: Mean of religion_importance
mean_religion_importance <- mean(survey_data_clean$religion_importance, na.rm = TRUE)
mean_religion_importance
```

## Explanation of `mutate()` Logic:

- **`mutate()`**: Adds or modifies columns in a data frame.
- **`ifelse()`**: Replaces values matching the condition (`%in% non_applicable`) with `NA`. Values not matching remain unchanged.
- **`%in%`**: Checks if a value exists in the `non_applicable` list.

With this setup, non-applicable values are now consistently handled as `NA`, making them compatible with R's analysis functions.
