\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{geometry}
\geometry{margin=1in}

\title{Quiz 3}
\author{POL 201 - POL 501}
\date{}

\begin{document}
\maketitle


As a campaign organizer, you are preparing to host a political event. The event is free for attendees, but the owners of the venue will charge you a variable cost of \$70 per attendee. Additionally, there is a fixed cost for the venue rental of \$20,000 you have to pay regardless of how many people attend the event. The number of attendees is represented by a random variable $A$. Below is the distribution for the number of attendees:\footnote{Note: For simplicity, in this quiz we'll assume that the total number of attendees will fall exactly on one of the values listed in the table. }

\begin{table}[h!]
\centering
\begin{tabular}{c|c}
    \hline 
    Attendees ($A$) & Probability ($P(A = a)$) \\
    \hline
    $A < 50$ & 0 \\
    50 & 0.10 \\
    100 & 0.20 \\
    150 & 0.30 \\
    200 & 0.25 \\
    250 & 0.15 \\
    $A > 250$ & 0 \\
    \hline 
\end{tabular}
\caption{Probability Mass Function of Attendees.}
\end{table}
\begin{itemize}
    \item \textbf{Fixed Cost:} The venue rental fee is \$20,000.
\end{itemize}


\subsection*{Questions}

\subsubsection*{(a) Calculate the Expected Value of the Number of Attendees (3 points)}
Determine the expected value of the number of attendees at the event. \emph{Justify your answer}.

\begin{center}
\fbox{\parbox{0.95\textwidth}{\vspace{5cm} \hspace{1cm}
}}
\end{center}

\subsubsection*{(b) Define the Total Cost as a Linear Combination (1 points)}
To organize the event, the venue is charging you a cost per attendee of \$70, which will be totaled when the event finishes. Write down the total cost of the event, $C$, as a linear combination of $A$, the cost per attendee, and the venue rental fee (Hint: express $C$ as a function of the number of attendees times the cost per attendee plus the venue rental fee). \emph{Justify your answer}.

\begin{center}
    \fbox{\parbox{0.95\textwidth}{\vspace{2cm} \hspace{1cm}}}
 \end{center}

\subsubsection*{(c) Compute the Probability Mass Function of the Total Cost (3 points)}
Compute the probability mass function of the total cost, $C$. In other words, for each possible value of the total cost indicate its probability, $P_C(c)$. Express this in a table. \emph{Justify your answer}.

\begin{center}
    \fbox{\parbox{0.95\textwidth}{\vspace{7cm} \hspace{1cm}}}
\end{center}

\subsubsection*{(d) Compute the Cumulative Distribution Function of the Total Cost (2 points)}
Compute the cumulative distribution function of the total cost, $C$. In other words, for each possible value of the total cost calculate $P_C(C\leq c)$. Express this in a table. \emph{Justify your answer}.

\begin{center}
    \fbox{\parbox{0.95\textwidth}{\vspace{7cm} \hspace{1cm}}}
\end{center}

\subsubsection*{(e) Compute the Expected Value of the Total Cost (3 points)}
Calculate the expected value of the total cost of the event. \emph{Justify your answer}.

\begin{center}
    \fbox{\parbox{0.95\textwidth}{\vspace{5cm} \hspace{1cm}}}
 \end{center}

\subsubsection*{(f) Assess Budget Constraints (2 points)}
Your supervisor has indicated that the maximum budget for the event is \$35,000. Calculate the probability that the cost of the event exceeds this budget. \emph{Justify your answer}.

\begin{center}
    \fbox{\parbox{0.95\textwidth}{\vspace{5cm} \hspace{1cm}}}
 \end{center}



\subsubsection*{(g) Determine the Deviation Range (1 point)}
Suppose you communicate to your supervisor the value of the expected cost of the event. She wonders whether the typical deviation from that cost is such that \$35,000 is outside or inside of that typical deviation range. 

\textbf{Hint:} To answer this question, calculate the standard deviation of the total cost and find the range defined by the expected value plus and minus one standard deviation ($\mu + \sigma$ and $\mu - \sigma$). Determine whether \$35,000 is inside this interval. \emph{Justify your answer}.

\begin{center}
    \fbox{\parbox{0.95\textwidth}{\vspace{5cm} \hspace{1cm}}}
 \end{center}

\end{document}

Solve this. Use Python for your calculations but plan before using Python so fix step by step and before doing Python strategize and plan response and then you just fight them to calculations and address each letter one at a 